### 2011-11-07 doc/pldi2012/highlevel.tex
high level

\section{High-Level Runtime} 
\label{sec:highlevel}
In this Section we describe the high-level runtime system
that is designed to implement the Legion programming
and execution models while remaining machine agnostic for portability.  
The high-level runtime system operates on top of the 
low-level runtime which hides machine specific details
and is described in Section \ref{sec:lowlevel}.

The biggest challenge in implementing the high-level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high-level runtime employs two useful 
scheduling techniques.  The first technique, validated in 
Section~\ref{sec:exec}, is to make all scheduling decisions for a
task locally, i.e. in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies by decoupling the issuing
of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime.

@@

A deferred execution model however presents challenges to implementing
the semantics of the Legion programming model.  Deferred operations
may not be executed in the order they were scheduled if no dependency
is expressed between them.  To orchestrate the execution of Legion
programs in this environment the high-level runtime uses an architecture
similar to that of a pipelined out-of-order hardware processor, but at
a coarser granularity.  Tasks can be though of as corresponding to instructions
and regions to registers.  The runtime uses a pipeline of task execution stages
to extract
as much task-level parallelism as possible from an application, while
preserving the guarantees of the Legion programming model.  To illustrate
the implementation of the high-level runtime we now describe the
details of how each stage of task execution is performed.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  Tasks are
registered with the high-level runtime in sequential program order.  The
registration lists the logical
regions to be accessed, specifying the needed privileges and the coherence.
The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context (i.e. called by the same parent task).  These checks will
be sufficient to ensure correct program order, as was proved in Section~\ref{sec:exec}.

@@

%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.
%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.  
\subsubsection{Dependencies From Coherence and Privileges}
\label{subsec:cohdep}
Conceptually, dependence detection with a previous task requires comparing the two sets of
logical regions being accessed.  For each logical reason from the previous task
that may overlap with a logical region needed by the new task, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
only read privileges are in use, there is never a conflict, but if either or
both task needs write or reduction privileges, the coherence modes are compared
according to this table:

@@

An entry of {\tt Dep} means there is always a dependence, whereas {\tt None}
means that there is no dependency created by this region pair.  An entry of
{\tt Same} indicates a dependency that can be avoided if the two tasks are
able to use the same physical region.  Finally, an entry labeled {\tt Cont}
indicates a dependency that can be avoided if a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task will
generally have to wait for the older task to complete before it can begin
execution.

@@

Although the above algorithm for dependence detection is correct, the
high-level runtime is able to perform the detection much more efficiently with
the help of a custom data structure.

To perform dependence detection, the runtime will first find the {\em root region} for the 
requested region.  The root region is the ancestor region for which the parent task
has priviliges.  Once the root region for the requested region is found, 
the runtime can walk down the region tree from the root region to the node
for the requested region.  Each node contains a list of tasks that are
actively using the region corresponding to the node, potential dependencies are
discovered by comparing privileges and coherence modes with any tasks on that
list.

After check the list of tasks actively using the current node, the runtime
checks to see if one of the partitions of current node is {\em open}.  An
open partition indicates that there are some active tasks using subregions
of the current region.  No more than one partition of a node can be open
at any given time.  If the open partition is is along the path to the desired
region's node, traversal continues.  If the open partition lies along another 
path, or if the current node is the intended destination, the runtime must
{\em close} the currently active partition because it cannot guarantee
disjointness with the region the new task wishes to access.  To close a
partition in the region tree, the runtime must find all active tasks in that
subtree and consider them all to be dependencies for the task currently being
registered.  With the previously active partition closed, the runtime can 
then open up the desired partition and traverse further down the tree as
needed.

When the runtime traverses an open partition, its behavior depends on the kind
of partition being traversed.  If the partition is disjoint, no other subregion
can overlap with the target region, so the state of the other subregions can
be ignored in the traversal.  However, if the partition is aliases, any one
of those other subregions could overlap with the current partition, so
the runtime must close all of the open subregions (recording new dependencies
as it does so) before opening the subregion it intends to.

@@

\subsubsection{Managing Data Versions}
\label{subsec:dataflow}
One aspect of the deffered execution model is that it allows dependence analysis to run
ahead of the actual execution of tasks.  As a result the runtime must maintain information
about which {\em physical instances} of a logical region a task is allowed to use when executing.
A physical instance is a copy of the data in a logical region.  There can be multiple
physical instances of a logical region.  It is the responsibility of the high-level runtime
to ensure tasks use instances corresponding to the appropriate version of data.

%In addition to computing the dependences between tasks, we also have to compute the version
%of data that a task is permitted to use when it goes to execute since the computation.
%It is insufficient to simply record the producer task as there
%may be multiple active tasks for a logical region, all of which may produce different physical
%instances.  In addition, it is likely that the producer tasks have not been mapped
%yet which makes it impossible to know the set of physical instances from which to pull data.

In an out-of-order processor this problem is solved by register-renaming which maps logical
registers to a larger set of physical registers.  Our solution employs a similar technique
that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
is a place holder for the set of physical instances which a task will reference.  We call these
abstract instances because we may not know the actual physical instances until a task's predecessors
are mapped as described in Section \ref{subsec:mapping}.  For each region in
a task we record the source and destination abstract instances that correspond to the physical
instances from which a task will read its input data and then place its output data.
%When the task is mapped as described in section \ref{subsec:mapping} we will update these
%abstract instances with the actual physical instances chosen.

To determine the source and destination abstract instances of a region, we keep track of 
the previous valid abstract instances that were observed while traversing the region tree
during dependence analysis.  Every region node in the region tree that has active tasks
has a valid abstract instance.  Every time we close a region tree we close all the valid
abstract instances.  When a node is reopened it must then create a new abstract instance
to reflect that new physical instances must be created.
%If a region
%subtree must be closed then all the abstract instances in that subtree are marked closed
%so no additional tasks can register them as sources.  We can also chose to create a new
%abstract instance to replace a currently valid abstract instance to avoid a WAR dependence.
%By creating an abstract instance we will force the mapper to choose a different physical
%instance from the existing ones when mapping is performed.

Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
instance is a node in the dataflow graph corresponding to a specific version of a logical 
region and its instantiations as physical regions.  Tasks are then edges between abstract
instances indicating operations on different versions of data.  By computing this information
during dependence detection, we can record it before tasks begin executing out of order allowing
us to guarantee program correctness.  
%Continuing
%the analogy with out-of-order processors, this implicit construction of the dataflow graph
%is annalogous to register renaming which is done in order, but allows the processor to remove
%WAR dependences and execute instructions in parallel that may have had false dependences.

\subsubsection{Circuit Execution Example}
\label{subsec:cirdependence}
To illustrate concretely how dependence analysis works, we describe the steps for
performing dependence analysis for circuit example from Section \ref{sec:ex}.  
For conciseness we will only show the node region tree.  At the start of the task, there exists
a single abstract instance at the root node representing the valid set of physical
instances from the parent task.

When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
analysis for two node regions.  In the {\tt private\_node} case they will first open
up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
for each of their subregions.  We use a bold circle on an edge to indicate that it has been
opened.  Since all of the private regions use a different subregion
of a disjoint partition for the private node accesses, no dependence is detected.

Similarly, when the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
each of the necessary partitions are opened.  There is no dependence with any of the private
nodes since all the ghost nodes traverse a different subregion of the disjoint partition
{\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
{\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
The state of the region tree after dependence analysis has been performed for the first
{\tt compute\_new\_current} task can be seen in figure~\ref{sfig:mapping_fig:cnc}.

In the second phase of the computation, all of the {\tt distribute\_charge} tasks
will be issued.  For these tasks they will first traverse the region tree
for the regions containing their private nodes.  The path to these regions are
already open, but when the dependence analysis arrives at the specific private node subregion it will
discover a dependence between the task and the {\tt compute\_new\_currents} task that operated
on that region previously.  This dependence will be recorded and the 
{\tt distribute\_charge} task will be made the new active task as it dominated the
previous task.

For the ghost node regions, the dependence analysis will discover that the
there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect this as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region tree after dependence analysis for first {\tt distribute\_charge} task can
be seen in figure~\ref{sfig:mapping_fig:dc}.

For the final phase of the computation, all of the {\tt update\_voltage} tasks
are issued requiring their private node region and their shared node region.  Again
the private node paths are already open, but a WAR dependence is detected which
requires a new abstract instance to be created to avoid a dependence.  The
shared nodes present a more interesting case.  The first shared node to be registered
will detect a dependence with every prior task since it must open the shared partition
instead of the ghost partition.  A new abstract instance is created at the 
{\tt all\_shared} node and copies are recorded for each valid instance in the ghost partition
back to the new {\tt all\_shared} abstract instance.  After this is complete, the
ghost partition can be opened.  Each corresponding {\tt update\_voltage} task will record
a dependence on the copy operation to close ghost partition as it traverses the {\tt all\_shared}
node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
tasks.  The state of the region tree after the first update voltage task
has been registered can be seen in figure \ref{sfig:mapping_fig:volt}.

\subsection{Mapping}
\label{subsec:mapping}
After tasks have undergone dependence analysis, the tasks are placed on either a ready
queue to be mapped or waiting queue.  A task is placed on the ready queue if all the tasks with
which it registered a dependence have mapped.  If any of the task's dependences have
not mapped then the task is placed on the waiting queue.  The task also registers
itself with the task on which it depends so that when that task is mapped it will notify
the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
their data dependences.  By mapping tasks based on data dependences we guarantee the 
semantics of the programming model, while still allowing tasks to execute out of order.

When the scheduler is invoked, it will pull a series of tasks off the ready queue, map them,
and then issue them to be executed by the low level runtime.  The process of mapping a task
consists of several different queries made by the runtime to an application level object
called a {\em mapper}.  A mapper is an object which has control over how both tasks and
data can be placed in the machine.  This idea is a dynamic instantiation of the idea
initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
tasks and data are placed in a memory hierarchy.  Rather than rely on a static mapping
we opted for a dynamic interface that could make decisions both based on application specific
input data as well as the machine specifics both at start-up and throughout the execution
of an application.  %The interface for a mapper is described in Section \ref{subsec:mapinter}.

\subsubsection{Mapper Interface}
The mapper interface is a collection of eight function calls that allow the runtime to query
the application for information about how to map tasks are regions.  Mapper calls are made
from the runtime to ask where physical instances of regions should be place in the memory hierarchy,
which physical instances should be the source for copy operations, on which processors
tasks should be run, and how task stealing should be performed.
%\label{subsec:mapinter}
%The interface for a mapper is designed to give the programmer control over how data and tasks
%are placed in the machine.  Every mapper must support the following function calls:

@@

These eight task calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface does give the programmer
complete control over data and task movement, it will never allow the programmer to affect the
correctness of the program.  The high-level runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.

In order to make it easy for programmers to write programs, we provide an initial implementation 
of the mapper interface that uses simple heuristics to place tasks and data.  For example, when
performing task stealing, the default mapper only encourages stealing tasks in groups that use
the same logical regions, which improves locality by moving tasks that use the same regions
together.  The default mapper implementation makes it easy for the programmer to get an 
initial program running.  As the programmer tunes the application, he can write mappers that
specialize for both application specific and/or machine specific parameters.

The high-level runtime also supports the use of multiple mappers in the same application.  For each task 
call the programmer can choose a mapper to be used for the runtime calls made inside that
task.  This allows for the creation of library code which provides its own mappers indpendent
from any application level mappers.

\subsubsection{Task Stealing for Load Balance}
\label{subsec:steal}
In order to achieve load balancing in an application the runtime supports an interface
for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
used by Cilk \cite{CILK95} with guidance from the mapper about how stealing should be
performed.  Tasks are initially placed on the ready queue on the processor in which they
were registered.  However, a task can be moved to a different processor either by a mapper
explicitly choosing to relocate it, or by a steal request from another processor's mappers.

@@

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give to the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set which
it is going to permit to be stolen if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we are giving the mappers complete control over where tasks are mapped.
In addition, we are also giving the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to explicit processors with no interference
from the runtime.

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high-level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions which the task requires.  It then issues the necessary copies required
for the task to execute contignent on the tasks producing the source physical instances
have finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task haveing finished as well as the the set of tasks
on which the issuing task dependends having finished.  These dependences are all encoded
using event dependencies in the low level runtime.  Events are described
in detail in Section \ref{subsec:events}.  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

@@

\subsection{Task Completion}
\label{subsec:cleanup}
We define a task to have {\tt completed} once all of the task's regions reflect
the updated state of the task and all its sub-tasks.  This reflects an underlying assumption 
of the mapping process that the regions for which a task has priviliges will represent the data produced 
by that task.  However, this is not inherently true that a physical instance of a task reflects
this state after the task finishes executing.  If a task has child tasks these tasks may
still be executing when the parent task finishes the execution phase.

In order to handle this scenario the runtime must compute the necessary copy operations
from child task physical instances back to the parent task's physical instances.  To
compute these copy operations, the runtime must first wait until all the child tasks
have been mapped.  The runtime then issues a close operation to the each of the parent
task's region trees.  These close operations record all of the necessary copies to
restore the parent task's regions.  We call these copy operations at the end of 
a task {\em restoring copies}.
The runtime then issues these copies with the required dependences
and marks the task completed once the restoring copy operations are complete.
This is similar to the write-back phase of an out-of-order processor where store instructions cannot be
considered complete until the store effects the state of memory.

>>>>

\section{High-Level Runtime} 
\label{sec:highlevel}
%In this Section we describe the high-level runtime system
%that is designed to implement the Legion programming
%and execution models while remaining machine agnostic for portability.  
%The high-level runtime system operates on top of the 
%low-level runtime which hides machine specific details
%and is described in Section \ref{sec:lowlevel}.

The major challenge in implementing the high-level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high-level runtime employs two useful 
scheduling techniques.  The first, validated in 
Section~\ref{sec:exec}, is to make all scheduling decisions for a
task locally, i.e. in the context in which the
task was created.  The second is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies by decoupling the issuing
of operations
from when they are performed.  This model is supported
by features in the low-level runtime (see Section~\ref{sec:lowlevel}).

@@

A problem with deferred execution is that deferred operations must
still be executed in the correct order if they are dependent.
To ensure correctness the high-level runtime uses an architecture
similar to a pipelined out-of-order hardware processor, but at
the coarser granularity of tasks and regions rather than instructions and registers.
The stages of the task execution pipeline seek to extract
as much task-level parallelism as possible from an application, while
preserving dependencies.  We describe some
details of the four pipeline stages: dependence analysis, mapping, execution, and
task completion.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  Within a task call, subtasks are
registered with the high-level runtime in sequential execution order.  The
registration lists the logical regions to be accessed with needed privileges and coherence.
The runtime uses this information to perform
dependence detection with each of the subtasks already registered
in the same context (i.e. called by the same parent task).  
%These checks will
%be sufficient to ensure correct program order, as was proved in Section~\ref{sec:exec}.

@@

%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.
%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.  
\subsubsection{Dependencies From Coherence and Privileges}
\label{subsec:cohdep}
Conceptually, dependence detection between a new task $t$ and a previous task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region from $t'$
that may overlap with a logical region needed by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  Read
privileges never cause conflicts, but if either 
task needs write or reduction privileges, the coherence modes are compared:

@@

An {\tt Dep} entry indicates a dependence, whereas {\tt None}
means no dependence.  A
{\tt Same} entry indicates a dependence unless the two tasks 
use the same {\em physical region}.  Finally, a {\tt Cont} entry
indicates a dependence unless a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task 
waits for the older task to complete before it begins execution.

The table lists coherence modes we have not yet discussed.  Atomic
allows tasks to be reordered as long as the result is
serializable.  Both simultaneous and relaxed allow other tasks to
execute at the same time.  The difference comes in the restriction of
what data must be observed in the different modes.  In Simultaneous
mode, a task must see all updates to the logical region made by other
tasks operating on the same region at the same time.  In Relaxed mode,
a task is ambivalent as to whether it observes any updates to the
logical region by other tasks executing at the same time.

@@

To detect dependencies efficiently, the high-level runtime maintains a
local region forest (recall Section~\ref{sec:ex}) for each task.  The
root regions are the task's region arguments, and the forest includes
all partitions and subregions created within the task.  To perform
dependence detection on a region $r$, the runtime first finds a root
region $r'$ such that $r \rleq r'$.  The runtime then walks the region
forest from $r'$ to $r$.  Each region $r''$ on the path contains a
list $L$ of registered tasks using $r''$, and potential dependencies
are discovered by comparing privileges and coherence modes with the
tasks in $L$.

The runtime next checks if a partition of $r''$ is {\em open}.  An
open partition indicates that there are active tasks using subregions
of $r''$.  No more than one partition of a node can be open
at any time.  If the open partition is is along the path to $r$,
traversal continues.  If the open partition lies along another 
path the runtime must {\em close} the currently active partition because it cannot guarantee
disjointness with the region the new task wishes to access.  To close a
partition, the runtime finds all active tasks in that
subtree and considers them all dependencies for the new task.  
With the previously active partition closed, the runtime can 
then opens the desired partition and traverses further down the tree as
needed.

When the runtime traverses an open partition, its behavior depends on the kind
of partition being traversed.  If the partition is disjoint, no other subregion
can overlap with the target region, so the state of other subregions is
ignored.  However, if the partition is not disjoint, any 
of subregion could overlap with the desired subregion, so
the runtime must close all open subregions (recording new dependencies
as it does so) before continuing the traversal.

@@

\subsubsection{Managing Data Versions}
\label{subsec:dataflow}

A {\em physical instance} is a copy of the data in a logical region.  Multiple
physical instances of a logical region may exist simultaneously, and it is the responsibility of the high-level runtime
to ensure tasks use the correct version.
In an out-of-order processor this problem is solved by register-renaming which maps logical
registers to a larger set of physical registers.  Our solution employs a similar technique
that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
is a place holder for the set of physical instances a task will reference when it actually runs.  
Abstract instances are needed because we may not know the actual physical instances a task will
use until its predecessors in the dependence graph are {\em mapped} 
(see Section \ref{subsec:mapping}).  
To determine the abstract instances of a region, we keep track of 
the previous valid abstract instances that were observed while traversing the region forest
during dependence analysis.  Every region in the region forest that has active tasks
has a valid abstract instance.  Every time we close a region we close all the valid
abstract instances.  When a region is reopened it must then create a new abstract instance
to reflect that new physical instances must be created (i.e., a fresh version of the data must be copied from
the parent region).  In addition, whenever a task writes a region, a new abstract instance is usually created
to represent the new version of the data.

%For each region in
%a task we record the source and destination abstract instances that correspond to the physical
%instances from which a task will read its input data and then place its output data.
%When the task is mapped as described in section \ref{subsec:mapping} we will update these
%abstract instances with the actual physical instances chosen.
%In addition to computing the dependences between tasks, we also have to compute the version
%of data that a task is permitted to use when it goes to execute since the computation.
%It is insufficient to simply record the producer task as there
%may be multiple active tasks for a logical region, all of which may produce different physical
%instances.  In addition, it is likely that the producer tasks have not been mapped
%yet which makes it impossible to know the set of physical instances from which to pull data.

%If a region
%subtree must be closed then all the abstract instances in that subtree are marked closed
%so no additional tasks can register them as sources.  We can also chose to create a new
%abstract instance to replace a currently valid abstract instance to avoid a WAR dependence.
%By creating an abstract instance we will force the mapper to choose a different physical
%instance from the existing ones when mapping is performed.

%Abstract instances correspond to a dynamic construction of the dataflow graph.  Each abstract
%instance is a node in the dataflow graph corresponding to a specific version of a logical 
%region and its instantiations as physical regions.  Tasks are then edges between abstract
%instances indicating operations on different versions of data.  By computing this information
%during dependence detection, we can record it before tasks begin executing out of order allowing
%us to guarantee program correctness.  
%Continuing
%the analogy with out-of-order processors, this implicit construction of the dataflow graph
%is annalogous to register renaming which is done in order, but allows the processor to remove
%WAR dependences and execute instructions in parallel that may have had false dependences.

\subsubsection{Circuit Execution Example}
\label{subsec:cirdependence}
To illustrate concretely how dependence analysis works, we describe the steps for
performing dependence analysis for the circuit simulation in Listing~\ref{lst:code_ex}.
For conciseness we will only show the node region forest.  At the start of the task, there exists
a single abstract instance at the root representing the valid set of physical
instances from the parent task. 
When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
analysis for two node regions.  In the {\tt private\_node} case they will first open
up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub-partition
for each of their subregions.  We use a bold circle on an edge to indicate that it has been
opened.  Since all of the private regions use a different subregion
of a disjoint partition for the private node accesses, no dependence is detected.

Similarly, when the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
each of the necessary partitions are opened.  There is no dependence with any of the private
nodes since all the ghost nodes traverse a different subregion of the disjoint partition
{\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
{\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
The state of the region forest after dependence analysis has been performed for the first
{\tt compute\_new\_current} task can be seen in figure~\ref{sfig:mapping_fig:cnc}.

In the second phase of the computation, the {\tt distribute\_charge} tasks
are issued.  For these tasks they will first traverse the region forest
for the regions containing their private nodes.  The path to these regions are
already open, but when the dependence analysis arrives at the specific private node subregion it 
discovers a dependence between the task and the {\tt compute\_new\_currents} task that operated
on that region previously.  This dependence is recorded and the 
{\tt distribute\_charge} task becomes the new active task.

For the ghost node regions, dependence analysis discovers 
a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect this as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region forest after dependence analysis for the first {\tt distribute\_charge} task can
be seen in figure~\ref{sfig:mapping_fig:dc}.

For the final phase of the computation, all of the {\tt update\_voltage} tasks
are issued requiring their private node region and their shared node region.  Again
the private node paths are already open, but a write-after-read dependence is detected which
requires a new abstract instance to be created to avoid a dependence.  The
shared nodes present a more interesting case.  The first shared node to be registered
detects a dependence with every prior task since it must open the shared partition
instead of the ghost partition.  A new abstract instance is created at the 
{\tt all\_shared} node and copies are recorded for each valid instance in the ghost partition
back to the new {\tt all\_shared} abstract instance.  After this is complete, the
ghost partition can be opened.  Each corresponding {\tt update\_voltage} task will record
a dependence on the copy operation to close ghost partition as it traverses the {\tt all\_shared}
node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
tasks.  The state of the region forest after the first update voltage task
has been registered can be seen in Figure~\ref{sfig:mapping_fig:volt}.

\subsection{Mapping}
\label{subsec:mapping}
After dependence analysis a task is placed on either a {\em ready queue} to be mapped or a 
The high-level runtime maintains the invariant that a task {\em maps} only after all tasks on which it depends have mapped. 
While some predecessor tasks have not mapped a task remains in the {\em waiting queue}; otherwise it is placed on the {\em ready queue}.

%Tasks are mapped observing the partial ordering created by 
%their data dependences.  By mapping tasks based on data dependences we guarantee the 
%semantics of the programming model, while still allowing tasks to execute out of order.

When the scheduler is invoked, it pulls tasks off the ready queue, maps them,
and then issue them to be executed by the low level runtime.  The process of mapping a task
consists of several different queries made by the runtime to an application level {\em mapper} object.
A mapper controls how tasks and data can be placed in the machine.  Mappers are a dynamic version of 
the static mapping files of \cite{Fatahalian06}.  
%Rather than rely on a static mapping
%we opted for a dynamic interface that could make decisions both based on application specific
%input data as well as the machine specifics both at start-up and throughout the execution
%of an application.  %The interface for a mapper is described in Section \ref{subsec:mapinter}.

%\subsubsection{Mapper Interface}
The mapper interface is a collection of eight function calls that allow the runtime to query
the application for information about how to map tasks and regions.  Mapper calls are made
from the runtime to ask where physical instances of regions should be placed in the memory hierarchy,
which physical instances should be the source for copy operations, on which processors
tasks should be run, and how task stealing should be performed.
%\label{subsec:mapinter}
%The interface for a mapper is designed to give the programmer control over how data and tasks
%are placed in the machine.  Every mapper must support the following function calls:

@@

These eight mapper calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface allows the programmer
to tune performance, it cannot affect correctness.  The high-level runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.
To make it easy for programmers to write programs, we also provide a default
 mapper that uses simple heuristics to place tasks and data.  
%For example, when
%performing task stealing, the default mapper only encourages stealing tasks in groups that use
%the same logical regions, which improves locality by moving tasks that use the same regions
%together.  The default mapper implementation makes it easy for the programmer to get an 
%initial program running.  As the programmer tunes the application, he can write mappers that
%specialize for both application specific and/or machine specific parameters.

The high-level runtime also supports the use of multiple mappers in the same application.  Each task 
call may (but does not need to) have an associated mapper,
%runtime  can choose a mapper to be used for the runtime calls made inside that
%task.  
which allows for the creation of library code that provides its own mappers indpendent
from any application level mappers.

\subsubsection{Task Stealing for Load Balance}
\label{subsec:steal}
To achieve load balancing in an application the runtime uses
task stealing.  Our task stealing algorithm is loosely based on the Cilk algorithm
\cite{CILK95} with guidance from the mapper about how stealing should be
performed.  Tasks are initially placed on the ready queue on the processor in which they
were registered.  However, a task can be moved to a different processor either by a mapper
explicitly choosing to relocate it, or by a steal request from another processor's mappers.

@@

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set which
it is going to permit to be stolen, if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we give mappers complete control over where tasks are mapped.
In addition, we also give the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to specific processors with no interference
from the runtime.

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high-level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions the task requires.  It then issues copy operations required
for the task to execute contignent on predecessor tasks producing the source physical instances
having finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task haveing finished as well as the the set of tasks
on which the issuing task depends having finished.  These dependences are all encoded
using {\em events} (see Section \ref{subsec:events}).  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

@@

\subsection{Task Completion}
\label{subsec:cleanup}
We define a task to have {\tt completed} once all of the task's regions reflect
the updated state of the task and all its sub-tasks.  
%This reflects an underlying assumption 
%of the mapping process that the regions for which a task has priviliges will represent the data produced 
%by that task.  However, this is not inherently true that a physical instance of a task reflects
%this state after the task finishes executing.  
It is possible for a task to terminate but to have subtasks that are still in the execution phase.
%If a task has child tasks these tasks may
%still be executing when the parent task finishes the execution phase.  
To handle this scenario the runtime must compute the necessary copy operations
from child task physical instances back to the parent task's physical instances.  To
compute these copy operations, the runtime must first wait until all the child tasks
have been mapped.  The runtime then issues a close operation to the each of the parent
task's region forests.  These close operations record all of the necessary copies to
restore the parent task's regions.  
%We call these copy operations at the end of  a task {\em restoring copies}.
The runtime issues these copies with the required dependences
and marks the task completed once the copy operations are complete.
This is similar to the write-back phase of an out-of-order processor where store instructions cannot be
considered complete until the store effects the state of memory.
